---
title: "R_Project_Hannah_Enck"
output: html_document
date: "2025-07-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Packages to load.
```{r message=FALSE, warning=FALSE}
library(readr)
library(words2number)
library(dplyr)
library(stringr)
library(tidyverse)
```


Read in data.
```{r message=FALSE, warning=FALSE}
survey <- read.csv("https://raw.githubusercontent.com/andreavillanes/Project_Data_2025/refs/heads/main/Get%20to%20know%20survey%202025.csv")
```


Create a new column called Birth_Month which has only the month portion of Birth_Month_Year.
```{r message=FALSE, warning=FALSE}
survey$Birth_Month = sapply(strsplit(survey$Birth_Month_Year, ","), "[", 1)
```


Create a new column called Birth_Year which has only the year portion of Birth_Month_Year.
```{r message=FALSE, warning=FALSE}
survey$Birth_Year = sapply(strsplit(survey$Birth_Month_Year, ","), "[", 2)
```


Standardize the values for Birth_Month and Birth_Year.
```{r message=FALSE, warning=FALSE}
survey$Birth_Year = to_number(survey$Birth_Year)

survey <- survey %>%
  mutate(Birth_Month = recode(Birth_Month,
                        "1" = "January",
                        "2" = "February",
                        "3" = "March",
                        "4" = "April",
                        "5" = "May",
                        "6" = "June",
                        "7" = "July",
                        "8" = "August",
                        "9" = "September",
                        "10" = "October",
                        "11" = "November",
                        "12" = "December")) %>%
  mutate(Birth_Month = str_to_title(Birth_Month))
```


Show the five most frequent birth months and the five least frequent birth months.
```{r message=FALSE, warning=FALSE}
monthfreqs = sort(table(survey$Birth_Month), decreasing = TRUE)

# 5 most frequent birth months
top5months = survey %>% 
  count(Birth_Month, sort = TRUE) %>%
  head(5)
print(top5months)

# 5 least frequent birth months
least5months = survey %>% 
  count(Birth_Month, sort = TRUE) %>%
  tail(5)
print(least5months)
```


Show the five most frequent birth years and the five least frequent birth years.
```{r message=FALSE, warning=FALSE}
yearfreqs = sort(table(survey$Birth_Year), decreasing = TRUE)

# 5 most frequent birth years
top5years = survey %>% 
  count(Birth_Year, sort = TRUE) %>%
  head(5)
print(top5years)

# 5 least frequent birth years
least5years = survey %>% 
  count(Birth_Year, sort = TRUE) %>%
  tail(5)
print(least5years)
```


Clean the State column to eliminate any non-US state and standardize the values for State. 
```{r}
# define what counts as a US state
state_names <- c("ALABAMA", "ALASKA", "ARIZONA", "ARKANSAS", "CALIFORNIA", "COLORADO", "CONNECTICUT", "DELAWARE", "FLORIDA", "GEORGIA", "HAWAII", "IDAHO", "ILLINOIS", "INDIANA", "IOWA", "KANSAS", "KENTUCKY", "LOUISIANA", "MAINE", "MARYLAND", "MASSACHUSETTS", "MICIGAN", "MINNESOTA", "MISSISSIPPI", "MISSOURI", "MONTANA", "NEBRASKA", "NEVADA", "NEW HAMPSHIRE", "NEW JERSEY", "NEW MEXICO", "NEW YORK", "NORTH CAROLINA", "NORTH DAKOTA", "OHIO", "OKLAHOMA", "OREGON", "PENNSYLVANIA", "RHODE ISLAND", "SOUTH CAROLINA", "SOUTH DAKOTA", "TENNESSEE", "TEXAS", "UTAH", "VERMONT", "VIRGINIA", "WASHINGTON", "WEST VIRGINIA", "WISCONSIN", "WYOMING")

state_abbreviations <- c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA","HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY")

# create a recode map to change abbreviations to state names written out
recode_states <- tibble(state_abbreviations, state_names)
recode_map <- setNames(recode_states$state_names, recode_states$state_abbreviations)

survey$State <- str_trim(survey$State)

survey <- survey %>%
  mutate(State = recode(State, !!!recode_map, .default=State))

survey <- survey %>%
  mutate(State = toupper(State)) %>%
  mutate(State = case_when(State %in% recode_map ~ recode(State, !!!recode_map),
                           TRUE ~ NA_character_)) %>%
  mutate(State = str_to_title(State))

```


Show the five most frequent states and the five least frequent states. 
```{r}
# 5 most frequent states
top5states = survey %>% 
  count(State, sort = TRUE) %>%
  head(5)
print(top5states)

# 5 least frequent states
least5states = survey %>% 
  count(State, sort = TRUE) %>%
  tail(5)
print(least5states)
```


Standardize the Country column. Then show the five most frequent countries and the five least frequent countries.
```{r}
survey <- survey %>%
  mutate(Country = case_when(
    str_to_upper(Country) %in% c("USA", "US", "UNITED STATES OF AMERICA", "UNITED STATES") ~ "United States", TRUE ~ Country))

# 5 most frequent countries
top5countries = survey %>% 
  count(Country, sort = TRUE) %>%
  head(5)
print(top5countries)

# 5 least frequent countries
least5countries = survey %>% 
  count(Country, sort = TRUE) %>%
  tail(5)
print(least5countries)
```


Clean the How_Hear_About_Program column by eliminating the number and any extra character portion in those responses. Show the five most frequent responses to this column.
```{r}
survey$How_Hear_About_Program <- str_remove(survey$How_Hear_About_Program, "^\\d+\\. ")

#5 most frequent responses
top5responses = survey %>% 
  count(How_Hear_About_Program, sort = TRUE) %>%
  head(5)
print(top5responses)
```


Transform the data from wide to long on column Languages by duplicating the values of the other columns. 
Use the comma as a separator for Languages. Show the five most frequent languages. 
```{r}
transformed <- survey %>%
  separate_rows(Languages, sep = ',') %>%
  mutate(Languages = str_trim(Languages)) %>%
  count(Languages, sort = TRUE) %>%
  head(5)
print(transformed)
```
